EBASE manuscript

### 2023-04-18 manu-draft.qmd
added JMA comments

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data. Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\auni$). Respiration is modeled with the single parameter $r$ ($\runi$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($Sc$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\buni$. As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).

@@

where all parameters are defined as in @eq-ebase, with the additional parameters $p$, $\Theta$, and $K$ (the rearation coefficient). The BASE approach provides options to estimate $p$ and $\Theta$ from the data or to set as fixed values. A temperature dependency, $T$, is also included where respiration and gas exchange vary with temperature. The changes to @eq-base for EBASE were necessary for estuarine application and included 1) a linear relationship of production to PAR ($aPAR$) by setting $p = 1$; 2) constant respiration by removing the temperature dependency; and 3) use of a gas exchange parameterization commonly used in coastal or oceanic settings that depends on wind speed and temperature [i.e., @Grace15 used a littoral parameterization]. The justification for these changes is as follows.

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were 2012 observations at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15 minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, --84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $r$, and $b$.  All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].  Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors. For each parameter, low and high values for the means and standard deviations were assessed as changes approximately equal to two orders of magnitude of the means and standard deviations of the known parameters of the synthetic time series. The only exception was the $b$ parameter where the low values created priors with a distribution close to zero and the high values were equivalent to a uniform prior distribution in the range of 0 to 0.504 (justification noted in @sec-ebasemod). Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series.  The only models where the NSE values were less than zero for dissolved oxygen were those where $\rho$ of the $r$ parameter was low.  However, substantial variation in NSE values was observed among all other parameters.  Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`).  Conversely, the $a$ parameter was reproduced the most poorly, with a median value across all optimization periods and priors less than zero (median NSE = `r prmall[['a']]`). By optimization period, median NSE values across all parameters decreased slightly as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`).  The top three ranked models were identical for the seven and thirty day models and similar for models using the one day optimization period.  Median NSE values across parameters for the top model in each optimization period were also similar (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). For the seven day and thirty day optimization periods, the NSE values for all parameters were greater than zero for the top models, where two of the parameters in the top model for the one day optimization period has NSE < 0. Most parameters regardless of optimization period or priors had NSE < 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where the $\rho$ values were high for each prior distribution, regardless of the values for $\mu$, i.e., those with sufficient flexibility to search a larger parameter space regardless of $\mu$ produced metabolic parameter estimates most similar to the known values from the synthetic time series.

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters.  Median NSE values across all models and each optimization period for the different subsets of the prior combinations are shown.  For example, the bottom left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\rho$ value for the $a$ prior distribution was low or high.  The size of the points indicate the inter-quartile range of NSE values for each subset.  As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\rho$ values for each parameter. Although most of the median value are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.  For example, large differences in the median NSE values are observed for the low and high values of $\rho$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values using a high value using the 7 and 30 day optimization periods.  A similar, but smaller effect is seen for the $\rho$ value of the prior for the $b$ parameter.  A conclusion similar to that for @fig-priorcomp is that large standard deviations in the prior distributions lead to more accurate estimates of metabolic parameters.  Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\rho$.  This suggests that EBASE model sensitivity is most influenced by $\rho$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.

The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameters estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.  However, the $b$ parameter was estimated with high variability for the one and seven day optimization periods (@fig-optex a, c, bottom row) and with a large positive bias for the 30 day optimization period (@fig-optex e, bottom row).  These results were not unexpected because it was impossible to evaluate NSE estimates using a fixed value in the synthetic results and the accuracy was not considered in identifying the best performing model. Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).  The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\rho$.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, with the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters. Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE were uninformed using the same large $\rho$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).  A one day model optimization period was used for EBASE for comparability with the Odum method.

The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism, gross production, respiration, and gas exchange. Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method. For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher metabolic estimates). Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each. Root mean square errors (RMSE) of each linear model were also evaluated. Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated, as expected. The intercept values were all significant, excluding production with the observed dissolved oxygen, suggesting some level of bias when comparing estimates between the methods. The slope estimates for respiration and gas exchange were significantly different for estimates from the observed dissolved oxygen, where all but respiration were signicantly different using the detided dissolved oxygen. The RMSE values of each regression comparison were generally lower in the detided results for all comparisons, excluding gas exchange which was higher with the detided dissolved oxygen compared to the observed. This latter comparison suggests that detiding the time series produces more comparable estimates between methods.

As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]a). This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified with estimates from the Odum method as negative values for production (P) and positive values for respiration (R), i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., $a$ and $r$ cannot be negative, Equations [-@eq-aprior] and [-@eq-rprior]). Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.

@@

![The default prior distributions for $a$, $r$, and $b$ used in EBASE. All priors are normal distributions and truncated to positive values. $b$ is also truncated to a maximum of 0.504.](figs/priorplot.png){#fig-priorplot fig-align="center"}

@@

![Synthetic time series for one year of continuous data at Apalachicola Bay. Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for production (P), respiration (R), and gas exchange (D). The $a$ parameter is shown as an areal value for comparison with the metabolic results.  The $b$ parameter is also fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ and is not shown.](figs/synapa.png){#fig-synapa}

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combinations of prior distributions and optimization period. The means and standard deviations are evaluated as low (L) or high (H) values (see text for details) and the optimization period is one day, seven days, or thirty days. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate the mean of the synthetic data is a better predictor than EBASE.  The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the median NSE value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Summaries of median Nash-Sutcliffe Efficiency values across all parameters for the comparisons of EBASE and synthetic model outputs separated by different combinations of prior values. Each result on the x-axis represents the median values for which the combination prior applies, e.g., all results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.  Point size represents the inter-quartile range of median NSE values for each result.](figs/priorsumcomp.png){#fig-priorsumcomp}

@@

![Comparisons of daily metabolic estimates from the EBASE and Odum methods for one year of continuous monitoring data at Apalachicola Bay. Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model. All values are areal as mmol O$_2$ m$^2$ d$^{-1}$. Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}

>>>>

where the terms on the right side of the equation are gross primary production (P or $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data. Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\auni$). Respiration is modeled with the single parameter $r$ ($\runi$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($Sc$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\buni$. As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).

@@

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were observations from the year 2012 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, --84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $r$, and $b$.  All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].  Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$)(Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors. For each parameter, low and high values for the means and standard deviations were assessed as changes approximately equal to two orders of magnitude of the means and standard deviations of the known parameters of the synthetic time series. The only exception was the $b$ parameter where the low values created priors with a distribution close to zero and the high values were equivalent to a uniform prior distribution in the range of 0 to 0.502 (justification noted in @sec-ebasemod). Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series.  The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $r$ parameter was low.  However, substantial variation in NSE values was observed among all other parameters.  Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`).  Conversely, the $a$ parameter was reproduced the most poorly, with a median value across all optimization periods and priors less than zero (median NSE = `r prmall[['a']]`). By optimization period, median NSE values across all parameters decreased slightly as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`).  The top three ranked models were identical for the seven and thirty day models and similar for models using the one day optimization period.  Median NSE values across parameters for the top model in each optimization period were also similar (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). For the seven day and thirty day optimization periods, the NSE values for all parameters were greater than zero for the top models, where two of the parameters in the top model for the one day optimization period has NSE < 0. Most parameters regardless of optimization period or priors had NSE < 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where the $\sigma$ values were high for each prior distribution, regardless of the values for $\mu$, i.e., those with sufficient flexibility to search a larger parameter space regardless of $\mu$ produced metabolic parameter estimates most similar to the known values from the synthetic time series.

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters.  Median NSE values across all models and each optimization period for the different subsets of the prior combinations are shown.  For example, the bottom left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\sigma$ value for the $a$ prior distribution was low or high.  The size of the points indicate the inter-quartile range of NSE values for each subset.  As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter. Although most of the median NSE values are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.  For example, large differences in the median NSE values are observed for low and high $\sigma$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values when using a high $\sigma$ value using the 7 and 30 day optimization periods.  A similar, but smaller effect is seen for the $\sigma$ value of the prior for the $b$ parameter.  A conclusion similar to that for @fig-priorcomp is that large standard deviations in the prior distributions lead to more accurate estimates of metabolic parameters.  Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\sigma$.  This suggests that EBASE model sensitivity is most influenced by $\sigma$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.

The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.  However, the $b$ parameter was estimated with high variability for the one and seven day optimization periods (@fig-optex a, c, bottom row) and with a large positive bias for the 30 day optimization period (@fig-optex e, bottom row).  These results were not unexpected because it was impossible to evaluate NSE estimates using a fixed value in the synthetic results and the accuracy was not considered in identifying the best performing model. Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).  The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\sigma$.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism, as well as alternative statistical approaches to estimating the parameters. Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE were uninformed using the same large $\sigma$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).  A one day model optimization period was used for EBASE for comparability with the Odum method.

The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism (NEM), gross primary production (P), ecosystem respiration (R), and gas exchange (D). Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and linear regression fits to evaluate potential biases in each method. For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher metabolic estimates). Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each. Root mean square errors (RMSE) of each linear model were also evaluated. Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all metabolic estimates, regardless of method or the input time series, were positive and significantly correlated, as expected. The intercept values were all significant, excluding P with the observed dissolved oxygen, suggesting some level of bias when comparing metabolic estimates between the methods. The slope estimates for R and D were significantly different for metabolic estimates from the observed dissolved oxygen, where all but R were significantly different using the detided dissolved oxygen. The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed. This result suggests that detiding the time series produces more comparable metabolic estimates between methods.

As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) than using the observed dissolved oxygen as input ([@fig-apacmpfig]a). This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified with estimates from the Odum method as negative values for P and positive values for R, i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative). Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.

@@

![The default prior distributions for $a$, $r$, and $b$ used in EBASE. All priors are normal distributions and truncated to positive values. $b$ is also truncated to a maximum of 0.502.](figs/priorplot.png){#fig-priorplot fig-align="center"}

@@

![Synthetic time series for one year of continuous data at Apalachicola Bay. Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for gross primary production (P), ecosystem respiration (R), and gas exchange (D). The $a$ parameter is shown as an areal value for comparison with the metabolic results.  The $b$ parameter is also fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ and is not shown.](figs/synapa.png){#fig-synapa}

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combinations of prior distributions and optimization period. The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period is one day, seven days, or thirty days. The metabolic results for dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate the mean of the synthetic data is a better predictor than EBASE.  The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the median NSE value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Summaries of median Nash-Sutcliffe Efficiency (NSE) values across all parameters for the comparisons of EBASE and synthetic model outputs separated by different combinations of prior values. Each result on the x-axis represents the median values for which the combination prior applies, e.g., all results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.  Point size represents the inter-quartile range of median NSE values for each result.](figs/priorsumcomp.png){#fig-priorsumcomp}

@@

![Comparisons of daily metabolic estimates from the EBASE and Odum methods for one year of continuous monitoring data at Apalachicola Bay. Results for net ecosystem metabolism (NEM), gross primary production (P), ecosystem respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model. All values are areal as mmol O$_2$ m$^2$ d$^{-1}$. Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}
